Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-05-31 paper/paper.qmd
braindump

@@

@@

Random forest models were used to examine whether OTU data collected on the day
of diagnosis could classify CDI cases as severe according to four different
definitions of severity.
We used the mikropml R package v1.5.0 [@topcuoglu_mikropml_2021] implemented in
a custom version of the mikropml Snakemake workflow [@sovacool_mikropml_2023]
for all steps of the machine learning analysis.
We have a full dataset which uses all samples available for each severity
definition, and an intersection dataset which consists of only the samples that
have all four definitions labelled.
The intersection dataset is the most fair for comparing model performance across
definitions, while the full dataset allows us to use as much data as possible
for model training and evaluation.
Datasets were preprocessed with the default options in mikropml to remove
features with near-zero variance and scale continuous features from -1 to 1. <!-- No features had missing values and no features were perfectly correlated -->
During preprocessing, `r preproc_ranges$min_removed` to `r preproc_ranges$max_removed`
features were removed due to having near-zero variance, resulting in datasets
ranging from `r preproc_ranges$min_feats` to `r preproc_ranges$max_feats`
depending on the severity definition.
We randomly split the data into an 80% training and 20% test set and repeated
this 100 times, followed by training models with 5-fold cross-validation.
Model performance was calculated on the test set using the area under the
receiver-operator characteristic curve (AUROC) and the area under the balanced
precision-recall curve (AUBPRC).
Permutation importance was then performed to determine which OTUs contributed
most to model performance.
We reported OTUs with a significant permutation test in at least 80 of the 100
models.

>>>>

Numerous studies indicate that the gut microbiome may play a role in _C. diff_ 
colonization, infection, and clearance.

@@

Two goals: investigate whether we can predict CDI severity based on OTU data to
inform how the gut microbiome may modulate severity (ML-based science: good performance implies something about underlying biology), and determine whether there
is potential clinical value in OTU-based models.

@@

Random forest models were used to examine whether OTU data collected on the day
of diagnosis could classify CDI cases as severe according to four different
definitions of severity.
We used the mikropml R package v1.5.0 [@topcuoglu_mikropml_2021] implemented in
a custom version of the mikropml Snakemake workflow [@sovacool_mikropml_2023]
for all steps of the machine learning analysis.
We have a full dataset which uses all samples available for each severity
definition, and an intersection dataset which consists of only the samples that
have all four definitions labelled.
The intersection dataset is the most fair for comparing model performance across
definitions, while the full dataset allows us to use as much data as possible
for model training and evaluation.
Datasets were preprocessed with the default options in mikropml to remove
features with near-zero variance and scale continuous features from -1 to 1. <!-- No features had missing values and no features were perfectly correlated -->
During preprocessing, `r preproc_ranges$min_removed` to `r preproc_ranges$max_removed`
features were removed due to having near-zero variance, resulting in datasets
ranging from `r preproc_ranges$min_feats` to `r preproc_ranges$max_feats`
depending on the severity definition.
We randomly split the data into an 80% training and 20% test set and repeated
this 100 times, followed by training models with 5-fold cross-validation.
Model performance was calculated on the test set using the area under the
receiver-operator characteristic curve (AUROC) and the area under the balanced
precision-recall curve (AUBPRC).
Permutation importance was then performed to determine which OTUs contributed
most to model performance.
We reported OTUs with a significant permutation test in at least 75 of the 100
models.
